An enhancer code compendium of early human neuronal and neural crest development

### 2025-04-08 manuscript.tex
Go through results again.

We generated human neural tube organoids (hNTOs) to model early human neuronal and neural crest development. To this
end we embedded single cells from two induced pluripotent stem cell (iPSC) lines (Sigma and BJ1) and one embryonic stem
cell line (ESC; H9) into separate synthetic poly-ethylene glycol (PEG) based extracellular matrices. These were
incubated for 11 days and we supplemented the medium with retinoic acid (RA) and Smoothened agonist (SAG) from day 3
until day 5 (\textbf{Fig. 1a}). These conditions were previously shown to generate multiple neuronal progenitor and
neural crest states\cite{abdelfattahActuationEnhancesPatterning2021}. To characterize the organoid-derived cell types to a greater level of detail at both
transcriptomic as well as chromatin accessibility level we performed 10x single-cell multiomics (combined scATAC-seq and
scRNA-seq) at four time points of differentiation (day 4-5, day 6-7, day 8-9 and day 10-11; \textbf{Fig. 1a}),
resulting in 22,862 high quality cells (\textbf{Fig. 1b}). Note, that similar transcriptomic and chromatin accessibility
states were obtained from all three cell lines (\textbf{Supplementary Fig. S1a-b}) demonstrating the robustness of the
organoid system.\par
To evaluate the biological relevance of the cell types and states obtained from the organoid, we also performed 10x single
cell multiome on the head of a single four weeks post conception human embryo, resulting in 25,785 high quality cells (
\textbf{Fig. 1c}). We annotated both the organoid and embryo cells based on the expression of marker genes
(\textbf{Supplementary Table S1}). In both systems we could identify neuronal progenitor cells, early differentiating neurons,
neural crest cells, facial mesenchyme cells and early differentiating peripheral neurons (\textbf{Fig. 1b-c}).
Additionally, in the organoid only we could identify a pluripotent stem cell population (\textbf{Fig. 1b}) and cell
types missing from the organoid include endothelial cells, perivascular macrophages (PVM) / microglia (MIC) and otic placode (\textbf{Fig. 1c}).\par
Next, we sub-clustered the neuronal progenitor, early neuron, and neural crest cell types (\textbf{Fig. 1d-f}). Both
organoid and embryo neuronal progenitors show a gradient of different cell states reflecting dorso-ventral patterning as
is evident from the expression of dorso-ventral marker TFs (REF): from ventral to dorsal \textit{ARX}, \textit{NKX2-2}, \textit{PAX6}, \textit{IRX3}, \textit{ZIC1} and \textit{GDF7} (\textbf{Fig. 1d}). In both the organoid and embryo system we obtained multiple early neuronal differentiation states.
A cluster that still expresses the neuronal progenitor marker \textit{SOX2}, cells that express the early neuronal induction
marker \textit{ASCL1} and distinct clusters marked by the expression of \textit{GAD1}, \textit{UNCX} and \textit{GATA2} (\textbf{Fig. 1e}). Finally,
focusing on the neural crest cells, in both the organoid and the embryo we obtained a population of pre-migratory neural
crest (\textit{GRHL2} expression and absence of \textit{ZEB2} expression), migratory neural crest (\textit{SOX10} and \textit{ZEB2} expression), facial
  mesenchyme cells (\textit{TWIST1} expression) and early differentiation peripheral neurons (\textit{NEUROD1} expression). In the embryo,
  but not in the organoid, we observed an additional population of cells marked by the expression of \textit{SOX10} and \textit{PAX2}
reflecting the otic placode (\textbf{Fig. 1f}).\par
Given that we performed single-cell multiome on the entire head of the four weeks post conception embryo we also
assessed the expression of anterior-posterior marker TFs. While the organoid cells had either an anterior forebrain
(\textit{SIX3}) or anterior spinal cord (\textit{HOXB3}) identity, in the embryo we observed a gradient of anterior-to-posterior identities (REF). Marked by the expression of \textit{SIX3}, \textit{EMX2}, \textit{FOXG1}, \textit{BARHL1}, \textit{EN1 }and \textit{HOXB3} (\textbf{Fig. 1g}).\par
Cell cycle arrest is an important hallmark of the start of neuronal differentiation (REF), for this reason we scored
both organoid and embryo cells for the expression of marker genes of the G1, G2M and S phase of the cell cycle and
quantified the fraction of cells in each phase stratified by the main cell types (\textbf{Fig. 1h}). From this it is
clear that both early differentiating peripheral neurons and to a greater extent early differentiating neurons of the
central nervous system are in cell cycle arrest, while the other cell types are still cycling (\textbf{Fig. 1h}).\par
Finally, to assess whether the organoid-derived cell types and states are also similar to the embryo-derived cell types
and states on a chromatin accessibility level we identified sets of co-accessible regions in the organoid model using
topic modeling. We focused on topics (i.e., sets of co-accessible regions) that were enriched for marker TF motifs and
calculated the average expression of marker genes for each cell group where these regions were co-accessible, for both
organoid and embryo cells. We observed similar patterns of marker gene expression for the same cell state (defined by
the topic) across the organoid and embryo (\textbf{Fig. 1i}) with an average Pearson correlation coefficient of 0.83
across identical states (\textbf{Supplementary Fig. S1c-d}).\par

@@

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) \textbf{g}, UMAP based on gene expression profiles of organoid (top; UMAP axis 1 and 2 shown) and embryo (bottom; UMAP
    axis 2 and 3 shown) for neuronal progenitor subcluster along with the expression of indicated genes in red-green-blue
    color scale. \textbf{h}, UMAP based on gene expression profiles in organoid (top left) and embryo (top right) colored by
    cell cycle phase identity and quantification of percentage of each cell for each cell cycle phase stratified by cell
    type in the organoid (bottom left) and embryo (bottom right). \textbf{i}, Average gene expression level in
    $log_{2}(CPM)$
    (dot color) and fraction of cells expressing the gene (dot size) per cell state, defined by topic modeling, for organoid
    cells (top) and embryo cells (middle) and motif enrichment level as normalized enrichment score (NES; dot color) and
    number of regions enriched for the motif (dot size) per topic (x-axis) and motif (y-axis; bottom). PVM/MIC: perivascular
    microphage/microglia.
    Color scales for RGB scatter plots are as follows (in Log(CPM)). 
    For organoid:
    ARX [0, 1.5],
    NKX2-2 [0, 1.5],
    PAX6 [0, 1.5],
    IRX3 [0, 1.0],
    ZIC1 [0, 1.5],
    GDF7 [0, 1.5],
    SIX3 [0, 1.5],
    EMX2 [0, 1.5],
    FOXG1 [0, 0.8],
    BARHL1 [0, 1.0],
    EN1 [0, 1.5],
    HOXB3 [0, 1.5],
    SOX2 [0, 1.5],
    ASCL1 [0, 1.5],
    ISL1 [0, 1.5],
    GAD1 [0, 0.8],
    UNXC [0, 0.8],
    GATA2 [0, 0.8],
    GRHL2 [0, 1.5],
    ZEB2 [0, 3.0],
    SOX10 [0, 0.8],
    NEUROD1 [0, 1.5],
    PAX2 [0, 1.5],
    TWIST1 [0, 1.5]
    For embryo:
    ARX [0, 1.5],
    NKX2-2 [0, 1.5],
    PAX6 [0, 1.5],
    IRX3 [0, 1.5],
    ZIC1 [0, 1.5],
    GDF7 [0, 1.5]
    SIX3 [0, 1.5],
    EMX2 [0, 1.5],
    FOXG1 [0, 0.8],
    BARHL1 [0, 1.5],
    EN1 [0, 1.5],
    HOXB3 [0, 1.5],
    SOX2 [0, 1.5],
    ASCL1 [0, 1.5],
    ISL1 [0, 1.0]
    GAD1 [0, 0.8],
    UNXC [0, 0.8],
    GATA2 [0, 0.8]
    GRHL2 [0, 1.5],
    ZEB2 [0, 3.0],
    SOX10 [0, 0.8]
    NEUROD1 [0, 1.5],
    PAX2 [0, 1.5],
    TWIST1 [0, 2.0]
  }
\end{figure}

@@

To gain insight into the sequence characteristics underlying cell type-specific activity of neural tube-, neural crest-
and derived cell type enhancers, we trained two sequence-to-function model convolutional neural networks:
"$DeepNeuralTube^o$" and "$DeepNeuralTube^e$". The models use DNA sequence of genomic regions as input and
classify the regions based on their cell type- or state-specific chromatin accessibility (in a multi-label multi-class
fashion; \textbf(Fig. 2a)). The classes of the models are defined based on topic modeling. To get a fine-grained
definition of cell states, we performed topic modeling separately for the subset of neuronal progenitors, neural crest
(incl. facial mesenchyme and peripheral neurons) and early differentiating neurons for both the organoid cells (to
train $DeepNeuralTube^o$) and embryo cells (to train $DeepNeuralTube^e$). To validate the ability of the models to
predict cell type-specific enhancer \textit{activity}, we tested their performance on enhancers from the VISTA enhancer
browser (\textbf{REF}). Relevant for this study, this database contains enhancers active in the neural tube (n=349) and
facial mesenchyme (n=143). Neural tube enhancers could be classified with an area under the receiver-operator curve
(auROC) of 0.75 and 0.80 using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2b}; and
area under the precision-recall curve (auPR) of 0.21 resp. 0.29 (\textbf{Supplementary Fig. S2a-b})), facial mesenchyme
enhancers with an auROC of 0.78 and 0.79 using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$
(\textbf{Fig. 2c}; and auPR of 0.17 for both models (\textbf{Supplementary Fig. S2a-b})). Of note, a fraction of neural tube
enhancers (24 \%) are predicted to be specific to pre-migratory neural crest by both models.\par
%TODO: C.M.: I think the original VISTA enhancers were annotated based on localization only, so it would make sense that premig. NC would be very hard to differentiate from normal neural tube ectoderm. Might be good to point out why this overlap exists?
Next, we assessed the explainability of the models. Concretely we tested whether the model learned known
TF binding motifs as features. For this, we performed motif enrichment analysis on all regions for each topic, to find
known motifs underlying the accessibility of each region. We
also
calculated gradients, for each model, on the top 1,000 regions (as predicted by the model) for each topic followed by
TF-MoDISco (\textbf{REF}) to find \textit{de novo} motifs. For both known motifs and \textit{de novo}
motifs we calculated pairwise similarities using TOMTOM (\textbf{REF}) followed by clustering, resulting in 56 motif
clusters (\textbf{Fig. 2d}). Of note, most clusters contain known-motifs enriched in both organoid regions and embryo
(\textbf{Fig. 2e}) regions with the exception of cluster 13 that is organoid specific (AP1 motif; \textbf{Supplementary Table S2}) and cluster
26 which is embryo specific (CG-rich motif; \textbf{Supplementary Table S2}). Most clusters (\textbf{Fig. 2f-g} and
\textbf{Supplementary Table S2}) contain at least one motif learned \textit{de novo} by the organoid and/or embryo model.
Thus, $DeepNeuralTube^o$ and $DeepNeuralTube^e$ use known-motifs as features to classify regions based
on cell type-specific chromatin accessibility.\par
As an illustrative example, we investigated an enhancer located in one of the introns of \textit{GLI3} (hs111.0; \textbf{Supplementary Fig. S2c}). This region has neural tube specific activity in E11.5 mouse embryos (\textbf{Fig. 2h}). Both
$DeepNeuralTube^o$ and $DeepNeuralTube^e$ explain this region similarly (Pearson correlation of x) and
highlight the importance of multiple SOX TFBS (\textbf{Fig. 2h}). A version of this enhancer where 5\% of the
nucleotides are changed (hs111.1) does not have neural tube specific enhancer activity and both
models predict that this loss of activity is due to the loss of a sinlge SOX TFBS (\textbf{Fig. 2h}).\par
Finally, to also investigate the performance of the models on migratory neural crest, a cell type for which there are no
validated enhancers in the Vista enhancer browser, we selected a genomic region that is specifically accessible in the
migratory neural crest cell type of the organoid (\textbf{Supplementary Fig. S2d}). Both models predict this region
to be specific to the migratory neural crest class (\textbf{Supplementary Fig. S2e}) and nucleotide importance scores
reveal TFBS for a SOX monomer, a SOX dimer and AP2 (\textbf{Fig. 2i}). We tested the enhancer activity of this region
using a reporter assay in chicken embryos. Indeed the enhancer was specifically active in migratory neural crest 24
hours after injection in an HH4 embryo (\textbf{Fig. 2i}).\par

@@

\subsection*{
  The enhancer code underlying dorso-ventral patterning of neuronal progenitors
}
Focusing on neuronal progenitors, we identified topics representing cells and regions of different dorso-ventral
identities in both the organoid and the embryo data (\textbf{Fig. 3a}). From ventral to dorsal this is topic 36, 38, 33,
54 and 48 in the organoid data and topic 34, 38, 79, 88 and 58 in the embryo data. As an example we first investigated
two enhancers of \textit{SHH} that are strongly conserved and specific to the floorplate [\textbf{REF}]; SFPE1 and SFPE2.
\textit{SHH} is most strongly expressed in the most ventral topic (33 and 34 resp. in the organoids and the embryo; \textbf{Fig. 3b}).
SFPE1 is located upstream of \textit{SHH} and SFPE2 in its second intron. Both enhancers are indeed most accessible in the cells
belonging to the ventral topic, in both the organoid and embryo data (\textbf{Fig. 3c}). Explaining the accessibility of both
enhancers using DeepNeuralTube reveals two FOX (one partial) and two RFX (one partial) binding sites in SFPE1
and two FOX, a ZEB and a TEAD binding site in SFPE2 (\textbf{Fig. 3d}).\par

To investigate the enhancer code underlying dorso-ventral patterning of neuronal progenitors further, we calculated
contribution scores on the top 1,000 regions of each organoid and embryo dorso-ventral topic using respectively $DeepNeuralTube^o$ and $DeepNeuralTube^e$. We then used TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018} to find reccurring seqlets for each topic (patterns) and we clustered patterns based on pairwise motif similarities. Next, we calculated the average contribution for all identified instances per cluster and for each topic resulting in a dorso-ventral enhancer code-table (\textbf{Fig. 3e}). Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns
underlying the different dorso-ventral states and the importance of the instances of these patterns is highly correlated across both
models (0.75 average Pearson correlation; \textbf{Fig. 3e}). For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD
and RFX instances is important (\textbf{Fig. 3e-f}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo cells (\textbf{Fig. 3g} and Supplementary Fig. S3). In subsequent more dorsal neuronal progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3e- g)}); PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3e-g} and Supplementary Fig. S3) and ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3e-g} and Supplementary Fig. S3). For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), SOX instances are important which correlates with \textit{SOX2/3} expression (\textbf{Fig. 3e-g} and Supplementary Fig. S3). Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3e} and Supplementary Fig. S3). Finally, we calculated the Jaccard index of genomic regions based on the presence of an instance of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig. 3h}). FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns tend to be present in separate regions (\textbf{Fig. 3h}).\par

It has been reported that FOXA1/2 is a pioneer factor (REF). Indeed, when centering on FOX instances in both the organoid and
embryo data we observe a strong footprint when looking at Tn5 cut-sites (\textbf{Fig. 3i}). We then
asked if there is any difference between regions that are generally bound by FOXA2 in any cell state that expresses this
TF and regions, bound by FOXA2, that are specific to the floorplate. For this purpose, we downloaded FOXA2 ChIP-seq
peaks in HepG2 and A549 and intersected these with floorplate peaks containing FOX instances. The floorplate peaks
that overlap with both ChIP-seq datasets we classified as regions that are generally bound by FOXA2 and those that
did not we classified as floorplate specific. Next, we quantified both the number of FOX instances per region and the
affinity of each individual instance (making use of the FOX motif score calculated using
FIMO\cite{grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025}) hypothesizing that the overall affinity of regions that are generally
bound by FOXA2 is higher than floorplate specific regions. Indeed, FOXA2 target regions that are specific to the floorplate mostly
have a single FOX instance while FOXA2 target regions that are general tend to have more instances (\textbf{Fig. 3j}).
Furthermore, in case there is a single FOX instance in a general FOXA2 target region the affinity of this instance tends
to be higher (\textbf{Fig. 3k}). Generating a \textit{de novo} motif from general and specific FOXA2 target regions that
have a single instance, we observe that FOX instances in regions that are general tend to be of the form
\textit{TGTTT\underline{AC}} while those instances in specific regions tend to be of the form \textit{TGTTT} (\textbf{Fig. 3l}).\par

@@

Given that we observed cell states from different AP domains in the neuronal progenitors from the embryo (but not
organoid; see \textbf{Fig. 1}) we decided to study the gene regulatory code underlying this patterning further. We
identified seven topics corresponding to different AP domains, from anterior to posterior this is Topic 61, Topic 59,
Topic 31, Topic 62, Topic 70, Topic 52 and Topic 71 (\textbf{Fig. 4a}). We first asked if the AP-program is distinct
from the DV program. In other words, is AP patterning encoded in the same or in different genomic regions as DV patterning? For this purpose, we calculated the pairwise correlation between region-topic
probabilities of all AP and DV topics (\textbf{Fig. 4b}). Both AP and DV patterning as well
as each of their sub-domains are mostly encoded in distinct genomic regions, with the exception of AP Topic 59 and 31
that correlate with the region-topic probability of DV Topic 38 (which corresponds to cells expressing \textit{NKX2-1/2}; \textbf{Fig. 4b})\par

>>>>

We generated human neural tube organoids to model early human neuronal and neural crest development. 
To this end we embedded single cells from two induced pluripotent stem cell (iPSC) lines (Sigma and BJ1) and one embryonic stem cell line (ESC; H9) into separate synthetic poly-ethylene glycol (PEG) based extracellular matrices. The organoids were incubated for 11 days and we supplemented the medium with retinoic acid (RA) and Smoothened agonist (SAG) from day 3 until day 5 (\textbf{Fig. 1a}). 
These conditions were previously shown to generate multiple neuronal progenitor and
neural crest states\cite{abdelfattahActuationEnhancesPatterning2021}. 
To characterize the organoid-derived cell types to a greater level of detail at both transcriptomic as well as chromatin accessibility level we performed 10x single-cell multiomics (combined scATAC-seq and
scRNA-seq) 
at four time points of differentiation (day 4-5, day 6-7, day 8-9 and day 10-11; \textbf{Fig. 1a}), 
resulting in 22,862 high quality cells (\textbf{Fig. 1b}). 
Note, that similar transcriptomic states were obtained from all three cell lines (\textbf{Supplementary Fig. S1a-b}) demonstrating the robustness of the organoid system.\par

To evaluate the biological relevance of the cell types and states obtained from the organoid, 
we also performed 10x single cell multiome on the head of a single four weeks post conception human embryo, 
resulting in 25,785 high quality cells (\textbf{Fig. 1c}). 
We annotated both the organoid and embryo cells based on the expression of marker genes
(\textbf{Supplementary Table S1}). 
In both systems we could identify neuronal progenitos, early differentiating neurons,
neural crest, facial mesenchyme and early differentiating peripheral neurons (\textbf{Fig. 1b-c}).
Additionally, in the organoid only we could identify a pluripotent stem cell population (\textbf{Fig. 1b}) 
while perivascular macrophages (PVM) / microglia (MIC) and otic placode cells were missing (\textbf{Fig. 1c}).\par

Next, we sub-clustered the neuronal progenitor, early neuron, and neural crest cell types (\textbf{Fig. 1d-f}). 
Both organoid and embryo neuronal progenitors show a gradient of different cell states reflecting dorso-ventral patterning, evident from the expression of dorso-ventral marker TFs\cite{delasRepressiveInteractionsGene2020}:
\textit{ARX}, \textit{NKX2-2}, \textit{PAX6}, \textit{IRX3}, \textit{ZIC1} and \textit{GDF7} (\textbf{Fig. 1d}). 
In both the organoid and embryo system we obtained multiple early neuronal differentiation states.
This includes a cluster that expresses the neuronal progenitor marker \textit{SOX2}, cells that express the early neuronal induction marker \textit{ASCL1} and distinct clusters marked by the expression of \textit{GAD1}, \textit{UNCX} and \textit{GATA2} (\textbf{Fig. 1e}). 
Finally, focusing on the neural crest, in both the organoid and the embryo we identified a population of pre-migratory neural crest (\textit{GRHL2} expression and absence of \textit{ZEB2} expression), migratory neural crest (\textit{SOX10} and \textit{ZEB2} expression), facial mesenchyme (\textit{TWIST1} expression) and early differentiation peripheral neurons (\textit{NEUROD1} expression). 
In the embryo, but not in the organoid, we observed an additional population of cells marked by the expression of \textit{SOX10} and \textit{PAX2} reflecting the otic placode (\textbf{Fig. 1f}).\par

Next, we assessed the expression of anterior-posterior marker TFs. 
While the organoid cells had either an anterior forebrain(\textit{SIX3}) or anterior spinal cord (\textit{HOXB3}) identity, 
in the embryo we observed a gradient of anterior-to-posterior identities. Marked by the expression of \textit{SIX3}, \textit{EMX2}, \textit{FOXG1}, \textit{BARHL1}, \textit{EN1 }and \textit{HOXB3} (\textbf{Fig. 1g})\cite{braunComprehensiveCellAtlas2023}.\par

Scoring cells of both organoid and embryo origin for the expression of cell cylce marker genes, 
we observed that both early differentiating peripheral neurons and to a greater extent early differentiating neurons of the central nervous system are in cell cycle arrest (\textbf{Fig. 1h}), an important hallmark of neuronal differentiating\cite{
braunComprehensiveCellAtlas2023}, while other cell types are still in different phases of the cell cycle (\textbf{Fig. 1h})\par

Finally, to assess whether the organoid-derived cell types and states are also similar to the embryo on a chromatin accessibility level 
we identified sets of co-accessible regions in the organoid model using topic modeling. 
Focusing on topics that were enriched for marker TF motifs we calculated the average expression of marker genes for each cell group where these regions were co-accessible for both organoid and embryo cells. 
We observed similar patterns of marker gene expression for the same cell state (defined by
the topic) across the organoid and embryo (\textbf{Fig. 1i}) 
with an average Pearson correlation coefficient of 0.83 across identical states (\textbf{Supplementary Fig. S1c-d}).\par

@@

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) \textbf{g}, UMAP based on gene expression profiles of organoid (top; UMAP axis 1 and 2 shown) and embryo (bottom; UMAP
    axis 2 and 3 shown) for neuronal progenitor subcluster along with the expression of indicated genes in red-green-blue
    color scale. \textbf{h}, UMAP based on gene expression profiles in organoid (top left) and embryo (top right) colored by
    cell cycle phase identity and quantification of percentage of each cell for each cell cycle phase stratified by cell
    type in the organoid (bottom left) and embryo (bottom right). \textbf{i}, Average gene expression level in
    $log_{2}(CPM)$
    (dot color) and fraction of cells expressing the gene (dot size) per cell state, defined by topic modeling, for organoid
    cells (top) and embryo cells (middle) and motif enrichment level as normalized enrichment score (NES; dot color) and
    number of regions enriched for the motif (dot size) per topic (x-axis) and motif (y-axis; bottom). PVM/MIC: perivascular
    microphage/microglia.
    Color scales for RGB scatter plots are as follows (in Log(CPM)). 
    For organoid: 
    Minimum is 0 and maximum is 1.5 Log(CPM) for all genes except:
    IRX3, BAHRL1: [0, 1.0], 
    FOXG1, GAD1, UNXC, GATA2, SOX10: [0, 0.8] and 
    ZEB2 [0, 3.0], 
    For embryo: 
    Minimum is 0 and maximum is 1.5 Log(CPM) for all genes except:
    FOXG1, GAD1, UNCX, GATA2, SOX10 [0, 0.8], 
    ISL1 [0, 1.0], 
    ZEB2 [0, 3.0] and
    TWIST1 [0, 2.0].
  }
\end{figure}

@@

In order to gain insight into the sequence characteristics underlying cell type-specific activity of neural tube-, 
neural crest- and derived cell type enhancers, we trained two sequence-to-function model convolutional neural networks:
"$DeepNeuralTube^o$" and "$DeepNeuralTube^e$". 
The models use DNA sequence of genomic regions as input and classify regions based on their cell type- or state-specific chromatin accessibility (in a multi-label multi-class fashion; \textbf(Fig. 2a)). 
The classes of the models are defined based on topic modeling. 
To get a fine-grained definition of cell states, we performed topic modeling separately for the subset of neuronal progenitors, neural crest (incl. facial mesenchyme and peripheral neurons) and early differentiating neurons for both organoid (to train $DeepNeuralTube^o$) embryo (to train $DeepNeuralTube^e$).\par

We tested the performance of $DeepNeuralTube^o$ and $DeepNeuralTube^e$ to predict cell type-specific activity 
using data from the VISTA enhancer browser\cite{viselVISTAEnhancerBrowsera2007, kosickiVISTAEnhancerBrowser2025}.
Enhancers active in the neural tube (n=349) could be classified with an area under the receiver-operator curve
(auROC) of 0.75 and 0.80 using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2b}; and
area under the precision-recall curve (auPR) of 0.21 resp. 0.29 (\textbf{Supplementary Fig. S2a-b})). 
Enhancers active in the facial mesenchyme (n=143) could be classified with an auROC of 0.78 and 0.79 
using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2c}; 
and auPR of 0.17 for both models (\textbf{Supplementary Fig. S2a-b})). 
Of note, a fraction of neural tube enhancers (24 \%) are predicted to be specific to pre-migratory neural crest by
both models. Given that the VISTA enhancers are annotated based on localization of enhancer activity only, and given
that is is difficult to visually distinguish between the neural tube and pre-migratory neural crest, this result 
is not surprising.\par

Next, we assessed the explainability of the models. Concretely we tested if the model learned known
TF binding motifs as features. To this end, we calculated pairwise similarities\cite{
  guptaQuantifyingSimilarityMotifs2007} between \textit{de novo} motifs learned by the model, 
and identified using TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018}, and known motifs that are enriched in cell type specific regions.
Clustering this pairwise similarity matrix resulted in 56 clusters (\textbf{Fig. 2d}). 
Most clusters contain known motifs enriched in regions specific to both organoid and embryo cell types (\textbf{Fig. 2e}), with the exception of cluster 13 that is organoid specific (AP1 motif; \textbf{Supplementary Table S2}) 
and cluster 26 that is embryo specific (CG-rich motif; \textbf{Supplementary Table S2}). 
More importantly, Most clusters contain at least one motif learned \textit{de novo} by the organoid and/or embryo model (\textbf{Fig. 2f-g} and \textbf{Supplementary Table S2}). 
%TODO: add percentages
Thus, $DeepNeuralTube^o$ and $DeepNeuralTube^e$ use known-motifs as features to classify regions based
on cell type-specific chromatin accessibility.\par

As an illustrative example, we investigated an enhancer located in one of the introns of \textit{GLI3} (hs111.0; \textbf{Supplementary Fig. S2c}). 
This region has neural tube specific activity in E11.5 mouse embryos (\textbf{Fig. 2h}). 
Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ explain this region similarly (Pearson correlation of x) 
%TODO: Add correlation
and highlight the importance of multiple SOX TFBS (\textbf{Fig. 2h}). 
A version of this enhancer where 5\% of the nucleotides are changed (hs111.1) does not have neural tube specific enhancer activity and both models predict that this loss of activity is due to the loss of a sinlge SOX TFBS (\textbf{Fig. 2h}).\par

Finally, to also investigate the performance of the models on migratory neural crest, 
a cell type for which there are no validated enhancers in the VISTA enhancer browser, 
we selected a genomic region that is specifically accessible in the migratory neural crest (\textbf{Supplementary Fig. S2d}). 
Both models predict this region to be specific to the migratory neural crest class (\textbf{Supplementary Fig. S2e})
and nucleotide importance scores reveal TFBS for a SOX monomer, a SOX dimer and AP2 (\textbf{Fig. 2i}). 
Next, we tested the enhancer activity of this region using an enhancer-reporter assay in chicken embryos. 
Indeed the enhancer was specifically active in migratory neural crest 24 hours after injection in an HH4 embryo (\textbf{Fig. 2i}).\par

@@

\subsection*{
  The enhancer code underlying dorso-ventral patterning of neuronal progenitors
}

Focusing on neuronal progenitors, we identified topics representing cells and regions of different dorso-ventral
identities in both the organoid and the embryo data (\textbf{Fig. 3a}). 
From ventral to dorsal this is topic 36, 38, 33, 54 and 48 for the organoids and
topic 34, 38, 79, 88 and 58 for the embryo. 
As an example we first investigated two enhancers of \textit{SHH} that are strongly conserved and specific to the 
floorplate: SFPE1 and SFPE2\cite{jeongDistinctRegulatorsShh2003}. 
\textit{SHH} is most strongly expressed in the most ventral topic (33 and 34 resp. for the organoids and the embryo; \textbf{Fig. 3b}).
SFPE1 is located upstream of \textit{SHH} and SFPE2 in its second intron. Both enhancers are indeed most accessible in the cells belonging to the ventral topic (\textbf{Fig. 3c}). 
Explaining the accessibility of both enhancers using DeepNeuralTube reveals two FOX (one partial) and two RFX (one partial) binding sites in SFPE1 and two FOX, a ZEB and a TEAD binding site in SFPE2 (\textbf{Fig. 3d}).\par

To investigate the enhancer code underlying dorso-ventral patterning of neuronal progenitors further, we clustered 
and aligned \textit{de novo} motifs, identified in doros-ventral regions, based on pairwise motif similarities. 
Next, we caclulated the average contribution for all identified motif instances per cluster and for each topic 
resulting in a dorso-ventral enhancer code-table (\textbf{Fig. 3e})
Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns underlying the dorso-ventral states and the 
importance of the instances of these patterns is highly correlated across both models (0.75 average Pearson correlation; \textbf{Fig. 3e}). 
For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD and RFX instances 
are important (\textbf{Fig. 3e-f}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo (\textbf{Fig. 3g} and Supplementary Fig. S3). 
In subsequent more dorsal neuronal progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3e- g)}); 
PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3e-g} and Supplementary Fig. S3) and 
ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3e-g} and Supplementary Fig. S3). 
For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), 
SOX instances are important that correlates with the expression of\textit{SOX2/3} (\textbf{Fig. 3e-g} and Supplementary Fig. S3). 
Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3e} and Supplementary Fig. S3). 
Finally, we calculated the Jaccard index of genomic regions based on the presence of an instance of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig. 3h}). 
FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns tend to be present in separate regions (\textbf{Fig. 3h}).\par

It has been reported that FOXA1/2 is a pioneer factor\cite{iwafuchi-doiPioneerTranscriptionFactor2016, iwafuchiGeneNetworkTransitions2020}. 
Indeed, when centering on FOX instances we observe a strong Tn5 cute-site footprint (\textbf{Fig. 3i}). 
We then asked if there is any difference between regions that are generally bound by FOXA2 in any cell state that expresses this TF and regions, bound by FOXA2, that are specific to the floorplate. 
For this purpose, we intersected FOXA2 ChIP-seq peaks in HepG2 and A549  with floorplate peaks containing FOX instances. 
We classified the regions that overlap with any of the ChIP-seq datasets as regions that are generally bound by FOXA2 and those that did no as floorplate specific. Quantifying both the number of FOX instances per region and 
the affinity of each individual instance (making use of the FOX motif score calculated using
FIMO\cite{grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025}) reveals that floorplate 
specific FOXA2 target regions have fewer FOX instances compared to general FOXA2 target regions (\textbf{Fig. 3j}). 
Furthermore, in case there is a single FOX instance in a general FOXA2 target region the affinity of this instance 
tends to be higher (\textbf{Fig. 3k}). These results suggest that the overall affinity of general FOXA2 target regions 
is higher (either by having a greater number of binding sites or by having binding sites with higher affinity).
Generating a \textit{de novo} motif from general and specific FOXA2 target regions that have a single FOX instance 
reveals that FOX instances in regions that are generally bound tend to be of the form \textit{TGTTT\underline{AC}} 
while those instances in specific regions tend to be of the form \textit{TGTTT} (\textbf{Fig. 3l}).\par

@@

In the embryo we observed neuronal progenitors states of different anterior-posterior (AP) identities (\textbf{Fig. 1g}).
We identified seven topics corresponding to different AP domains, from anterior to posterior this is Topic 61, Topic 59, Topic 31, Topic 62, Topic 70, Topic 52 and Topic 71 (\textbf{Fig. 4a}). 
We first asked if the AP-program is distinct from the DV program. 
In other words, is AP patterning encoded in the same or in different genomic regions as DV patterning? 
For this purpose, we calculated the pairwise correlation between region-topic probabilities of all AP and DV topics (\textbf{Fig. 4b}). 
This revealed that both AP and DV patterning are mostly encoded in distinct genomic regions, 
with the exception of AP Topic 59 and 31 that correlate with the region-topic probability of DV Topic 38 
(which corresponds to cells expressing \textit{NKX2-1/2}; \textbf{Fig. 4b})\par
